---
title: "Causal Relationships in the Quantitative Social Sciences II"
author: "A. Jordan Nafa"
date: "September 20th, 2022"
institute: "University of North Texas"
subtitle: "PSCI 3300.003 Political Science Research Methods"
format: 
  revealjs:
    theme: blood
    highlight-style: monokai
    toc-depth: 2
    self-contained: true
    citations-hover: true
    code-link: true
    transition: "fade"
    code-block-bg: "#272822"
    logo: "../../../../inst/figures/psci-3300-dag-trimmed-hex.png"
    css: "../../../../assets/slide-style.css"
    include-in-header: "../../../../assets/math-colors.js"
linkcolor: "#FF3030"
width: 1360
height: 800
bibliography: "../../../../assets/references.bib"
csl: "../../../../assets/apsa.csl"
link-citations: yes
---

```{r, include=FALSE}
## Load required packages
pacman::p_load(
  "tidyverse",
  "dagitty",
  "ggraph",
  "ggdag",
  install = FALSE
  )

## Base theme for the figures
fig_theme <- theme_light(base_size = 24, base_family = "serif") +
  ## Settings sepcific to the reveal.js theme
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(color = "white"),
    plot.subtitle = element_text(color = "white"),
    plot.caption = element_text(color = "white", face = "italic"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white", face = "bold"),
    legend.title = element_text(color = "white", face = "bold"),
    legend.text = element_text(color = "white", face = "bold"),
    strip.text = element_text(face = "bold"),
    plot.caption.position = "plot",
    legend.position = "top"
  )

shorten_dag_arrows <- function(tidy_dag, proportion){
  # Update underlying ggdag object
  tidy_dag$data <- dplyr::mutate(tidy_dag$data, 
                                 xend = (1-proportion/2)*(xend - x) + x, 
                                 yend = (1-proportion/2)*(yend - y) + y,
                                 xstart = (1-proportion/2)*(x - xend) + xend,
                                 ystart = (1-proportion/2)*(y-yend) + yend)
  return(tidy_dag)
}
```

## Overview

::: incremental
-   Formalizing the logic of cause and effect

    -   Potential Outcomes [@Rubin1974; @Rubin2005]

    -   Directed Acyclic Graphs [@Greenland1999; @VanderWeele2007]

-   Formal and informal representations of theories of politics

    -   Connecting research question, data, and theory gives us substance

-   Important Deadlines

    -   Deadline to form groups for the final project is Friday, September 30th

        -   Groups work on a single paper together--there appears to have been some confusion about this

    -   Problem Set I is due Friday, September 30th
:::

$$
\definecolor{treat}{RGB}{27,208,213}
\definecolor{outcome}{RGB}{98,252,107}
\definecolor{baseconf}{RGB}{244,199,58}
\definecolor{covariates}{RGB}{178,26,1}
\definecolor{index}{RGB}{37,236,167}
\definecolor{timeid}{RGB}{244,101,22}
\definecolor{mu}{RGB}{71,119,239}
\definecolor{sigma}{RGB}{219,58,7}
\newcommand{normalcolor}{\color{white}}
\newcommand{treat}[1]{\color{treat} #1 \normalcolor}
\newcommand{resp}[1]{\color{outcome} #1 \normalcolor}
\newcommand{sample}[1]{\color{baseconf} #1 \normalcolor}
\newcommand{covar}[1]{\color{covariates} #1 \normalcolor}
\newcommand{obs}[1]{\color{index} #1 \normalcolor}
\newcommand{tim}[1]{\color{timeid} #1 \normalcolor}
\newcommand{mean}[1]{\color{mu} #1 \normalcolor}
\newcommand{vari}[1]{\color{sigma} #1 \normalcolor}
$$

## Review of Key Terms

::: incremental
-   **Correlation**: A measure of the extent to which two or more variables tend to occur together.

    -   If two variables $\treat{X}$ and $\resp{Y}$ tend to occur together or increase at the same rate, we would say they are **positively correlated**

    -   If the occurrence of $\treat{X}$ unrelated to $\resp{Y}$, we would say these two variables are **uncorrelated**

    -   If when $\treat{X}$ occurs we are less likely to observe $\resp{Y}$, we would say these two variables are **negatively correlated**

-   **Line of Best Fit**: A line that minimizes how far data points are from the line on average, based on some measure of distance from data to the line
:::

## Linear Correlation Example

We can estimate this correlation in R with a simple linear model and data from the Varieties of Democracy project's `{vdemdata}` package

```{r, echo=TRUE}
# Get the data we need from the vdemdata package
vdem_df <- vdemdata::vdem %>% 
  # We'll use just the year 2018 here for simplicity
  filter(year == 2018) %>%  
  # Transmute a subset of the data for plotting
  transmute(
    country_name, 
    polyarchy = v2x_polyarchy*10, 
    gender_equality = v2x_gender*10
    )

# Estimate the linear relationship
lm_democ_gender <- lm(polyarchy ~ gender_equality, data = vdem_df)

# Print a summary of the result
broom::tidy(lm_democ_gender)
```

## Linear Correlation Example

```{r continuous-correlation-fit, echo=FALSE, dpi=300, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center', cache=TRUE}
# Initiate the plot object
ggplot(vdem_df, aes(x = gender_equality, y = polyarchy)) +
  # Add the data points
  geom_point(
    aes(fill = country_name), 
    show.legend = FALSE, 
    shape = 21,
    size = 3
    ) +
  # Color for the fill
  scale_fill_manual(values = viridisLite::turbo(n = length(vdem_df$country_name))) +
  # Labels for the plot
  labs(
    x = "Women's Political Empowerment Index",
    y = "Procedural Democracy",
    title = latex2exp::TeX(r'($E\[Y_{i} | X_{i}\] = -3.03 + 1.12\times X_{i}$ or $E\[Y_{i} | X_{i}\] = \alpha + \beta_{1}X_{i}$)')
  ) +
  # Adjust the x axis scales
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Adjust the y axis scales
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Plot the linear regression line
  geom_abline(
    intercept = -3.03, 
    slope = 1.12,
    color = "white",
    lty = 2,
    size = 2
    ) +
  # Plot theme settings
  fig_theme
```

## Review of Key Terms

::: incremental
-   **Mean** $\mean{\mu}_{\treat{x}}$: $\frac{\sum_{\obs{i}=\obs{1}}^{\sample{n}} \treat{x}_{\obs{i}}}{\sample{N}}$

    -   The average value of a numeric variable

    -   **Deviation from the mean** describes the distance between an observation's value for some variable relative to its mean

-   **Variance** $\vari{\sigma}_{\treat{x}}^{2}$: $\frac{\sum_{\obs{i}=\obs{1}}^{\sample{n}} (\treat{x}_{\obs{i}} - \mean{\mu}_{\treat{x}})^{2}}{\sample{N}}$

    -   A measure of how much variation a variable exhibits. It is the average of the square of the deviations from the mean
:::

## Review of Key Terms

::: incremental
-   **Standard Deviation** $\vari{\sigma}_{\treat{x}}$: $\sqrt{\vari{\sigma}_{\treat{x}}^{2}}$

    -   Another measure of the variability in some variable. It is the square root of the variance and has the advantage of being on roughly the same scale as the the variable itself

-   **Covariance** $Cov_{\treat{x}, \resp{y}} = \frac{\sum_{\obs{i}=\obs{1}}^{\sample{n}} (\treat{x}_{\obs{i}} - \mean{\mu}_{\treat{x}})(\resp{y}_{\obs{i}} - \mean{\mu}_{\resp{y}})}{\sample{N}}$

    -   A measure of the correlation between two variables. It is calculated as the average of the product of the deviations from the mean

-   **Correlation Coefficient** $\rho_{\treat{x}, \resp{y}} = \frac{Cov_{\treat{x}, \resp{y}}}{\vari{\sigma}_{\treat{x}} \cdot \vari{\sigma}_{\resp{y}}}$

    -   Another measure of the (linear) correlation between two variables. It is the obtained by dividing the covariance by the product of the standard deviations
:::

## Causal Questions

::: incremental
-   How does $\treat{X}$ influence $\resp{Y}$?

    -   Good research questions are broad but also precise

    -   Your research question should clearly state your <font style="color:#00F2FF">treatment</font> and <font style="color: #62FC6B">outcome</font> of interest

-   How does the <font style="color:#00F2FF">prospect of inclusion</font> for historically dispossessed social groups influence <font style="color: #62FC6B">voter turnout</font>?

-   Do <font style="color:#00F2FF">state-level policies</font> restricting abortion care adversely impact <font style="color:#62FC6B">access to contraception</font>?

-   How does the onset of <font style="color: #00F2FF">conflict</font> impact a country's <font style="color: #62FC6B">economic growth</font>?

-   Does <font style="color: #00F2FF">American military presence</font> abroad prevent political <font style="color: #62FC6B">instability</font>?
:::

## Defining Causal Effects

::: incremental
-   **Potential Outcomes** formally encode counterfactuals

    -   $\resp{Y}_{\obs{i}}(\treat{X} = 1)$ is the outcome we would observe if unit $i$ was treated

    -   $\resp{Y}_{\obs{i}}(\treat{X} = 0)$ is the outcome we would observe if unit $i$ was untreated

-   **Consistency** assumption connects observed outcomes to potential outcomes

    -   $\resp{Y}_{\obs{i}} = \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}})$ means we observe the potential outcome for the observed treatment

-   Causal effect for unit $i$ is $\resp{Y}_{\obs{i}}(\treat{X} = 1) - \resp{Y}_{\obs{i}}(\treat{X} = 0)$

    -   Difference in the observed and potential outcomes
:::

## Fundamental Problem of Causal Inference

::: incremental
-   For each individual $\obs{i}$ we can only observe $\treat{X}_{\obs{i}} = 1$ or $\treat{X}_{\obs{i}} = 0$

-   It is logically impossible to directly observe $\resp{Y}_{\obs{i}}(\treat{X} = 1) - \resp{Y}_{\obs{i}}(\treat{X} = 0)$

-   We can also generalize this to non-binary treatments (i.e., categorical, continuous, multivariate)
:::

## Key Assumptions

::: incremental
-   **Causal Ordering**: $\treat{X}_{\obs{i}} \longrightarrow \resp{Y}_{\obs{i}}$

    -   No reverse causality or simultaneity

-   **Consistency**: $\resp{Y}_{\obs{i}} = \resp{Y}_{\obs{i}}(\treat{x})$ if $\treat{X}_{\obs{i}} = \treat{x}$

    -   There are no hidden versions of the treatment or treatment variance is irrelevant

-   **No Interference between Units**: $\resp{Y}_{\obs{i}}(\treat{X}_{1}, \treat{X}_{2}, \dots, \treat{X}_{\sample{n}}) = \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}})$

    -   No causal effect of other units' treatment on other units' outcomes

-   **Consistency** + **No Interference** = **SUTVA** (stable unit-treatment variation assumption)
:::

## Manipulation

::: incremental
-   $\resp{Y}_{\obs{i}}(\treat{x})$ is the value that $\resp{Y}$ would take under $\treat{X}_{\obs{i}}$ set to $\treat{x}$

    -   Common to use uppercase to denote a theoretical quantity and lowercase to denote a **realization**

-   In principle, for the causal effect of $\treat{X}_{\obs{i}}$ to be properly defined, the treatment should be manipulable at least in theory

    -   There can be no causation without manipulation

-   This can be complicated for certain immutable characteristics such as race, sex, etc.

    -   One option is to focus on places where we could, at least in theory, manipulate these characteristics
:::

## Estimands

::: incremental
-   In a perfect world, we could estimate observation-level causal effects

    -   $\resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 1) - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0)$

    -   In practice this requires **strong assumptions** and is virtually impossible to identify

-   Instead, we usually attempt to estimate the **average causal effect** (ACE) in some subset of the population

-   Our estimand connects theory to data

-   Every study must be able to answer the question "What is your estimand?" [@Lundberg2021]
:::

## Estimands

::: incremental
-   **Sample Average Treatment Effect (SATE)**

    $$
    \mathrm{SATE} = \frac{1}{\sample{n}}\sum_{\obs{i}=1}^{\sample{n}}[\resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 1) - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0)]\\
    = \frac{\sum_{\obs{i}=1}^{\sample{n}}[\resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 1) - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0)]}{\sample{n}}
    $$

    -   Average of the outcomes if everyone is treated versus if no one is treated

    -   Answers the question what "would happen if everyone received the treatment compared to no one receiving it?"
:::

## Estimands

::: incremental
-   **Sample Average Treatment Effect for the Treated (SATT)**

    $$
    \mathrm{SATT} = \frac{1}{\sample{n}_{1}}\sum_{\obs{i}=1}^{\sample{n}}\treat{X}_{\obs{i}}[\resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 1) - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0)]\\ 
    = \frac{1}{\sample{n}_{1}}\sum_{\obs{i}=1}^{\sample{n}}\treat{X}_{\obs{i}}[\resp{Y}_{\obs{i}} - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0)]\\
    = \frac{\sum_{\obs{i}=1}^{\sample{n}}[\resp{Y}_{\obs{i}} - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0)]}{\sample{n}_{1}}
    $$

-   Answers the question "What is the average effect of the treatment among those who received treatment?"
:::

## Population versus Sample Estimands

::: incremental
-   SATE and SATT are specific to a given study

    -   Theoretically, these estimands are well defined without the kind of magical thinking people usually rely on for inference about a population (e.g., repeated sampling)

    -   They are **finite sample** or **finite population** estimands

-   What if we want to make inferences about a larger population though?

    -   Assume units are a **random sample** from a large/infinite population and hedge our bets on the central limit theorem

    -   In many areas of political science and international relations, this is not a logically defensible assumption
:::

## Population Estimands

::: incremental
-   **Population Average Treatment Effects (PATE)**

    $$
    \mathrm{PATE} = \mathrm{E}[\resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 1) - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0)]
    $$

-   **Population Average Treatment Effect for the Treated (PATT)**

    $$
    \mathrm{PATT} = \mathrm{E}[\resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 1) - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0) | \treat{X}_{\obs{i}} = 1]
    $$

-   In the case of a non-random sample from an **apparent population** where repeated sampling is logically impossible, $\mathrm{PATE} = \mathrm{SATE}$ and $\mathrm{PATT} = \mathrm{SATT}$

    -   Arriving at uncertainty estimates is less straightforward since we can't rely on magical thinking [@Berk1995; @Western1994; @Gill2020]
:::

## Other Interesting Estimands

::: incremental
-   **Conditional Average Treatment Effect (CATE)**

    $$
    \mathrm{CATE} = \mathrm{E}[\resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 1) - \resp{Y}_{\obs{i}}(\treat{X}_{\obs{i}} = 0) | \treat{X}_{\obs{i}} = \treat{x}]
    $$

    -   We can use this to detect heterogenous effects for theory testing or targeting

-   We will discuss additional estimands for direct and indirect causal effects later in the course
:::

## Summary

::: incremental
-   Causal inference is about **counterfactuals**

-   **Potential outcomes** constitute a way of formally expressing counterfactual relationships

    -   There are many possible estimands of interest--any possible difference between potential outcomes

-   Estimands connect research question, theory, and data

    -   A quantitative study that cannot answer the question "What is your estimand?" is effectively useless

-   Goal of contemporary political science is to provide vague answers to precise questions
:::

## Where We're Headed

::: incremental
-   Theoretical relationships as Bayesian networks

    -   Types of bias

    -   Difference between direct and indirect effects

-   Applied example to make this all more concrete

    -   Replicating @Barnes2020a

-   Moving from causal question to causal theory

    -   If you have not yet **at least** identified a causal relationship between some $\treat{X}$ and $\resp{Y}$ for your paper, you are now behind in the course
:::

------------------------------------------------------------------------

```{r dag-1, echo=FALSE, dpi=600, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center', cache=TRUE}
## Define the coordinates for the complex DAG
complex_coords <- list(
  x = c(X_L2 = 0, X_L1 = 1, X = 2, X_T1 = 3, Y_L2 = 0.5,
        Y_L1 = 1.5, Y = 2.5, Z_L2 = 0, Z_L1 = 1, Z = 2, Z_T1 = 3),
  y = c(X_L2 = 1, X_L1 = 1, X = 1, X_T1 = 1, Y_L2 = 0.5, 
        Y_L1 = 0.5, Y = 0.5, Z_L2 = 0, Z_L1 = 0, Z = 0, Z_T1 = 0)
)

## Plotmath labels for the complex DAG
complex_labels <- list(
  X_L2 = "...", Z_L2 = "...",
  Y_L2 = "Y[i*t-2]", X_L1 = "X[i*t-1]", 
  Z_L1 = "Z[i*t-1]", Y_L1 = "Y[i*t-1]",
  X = "X[i*t]", Z = "Z[i*t]",
  Y = "Y[i*t]", X_T1 = "...", Z_T1 = "..."
)

## Creating a More Complex DAG using ggdag syntax
complex_dag <- dagify(
  Y_L2 ~ X_L2 + Z_L2,
  X_L1 ~ X_L2 + Z_L1 + Y_L2,
  Y_L1 ~ X_L1 + X_L2 + Z_L1,
  Z_L1 ~ Y_L2 + Z_L2,
  X ~ X_L1 + Z + Y_L1,
  Y ~ X_L1 + Z + X,
  Z ~ Y_L1 + Z_L1,
  X_T1 ~ X + Y,
  Z_T1 ~ Z + Y,
  coords = complex_coords,
  labels = complex_labels
)

# Modifications for the contemporaneous effect of X on Y
complex_dag_contemp <- complex_dag %>%
  # Convert the DAG to a tibble
  tidy_dagitty() %>% 
  # Get the adjustment Set
  dag_adjustment_sets(exposure = "X", outcome = "Y") %>% 
  # Create Path-Specific colors and transparencies
  mutate(
    # Transparency for the edges
    .edge_alpha = case_when(
      name %in% c("X_L1", "Z", "X") & to %in% c("Y", "X") ~ 1,
      TRUE ~ 0.25
    ),
    # Color for the edges
    .edge_colour = case_when(
      name == "X" & to == "Y" ~ "#00FFFF",
      TRUE ~ "white"
    )
)

# Data for the plot annotations
period_labels <- tribble(
  ~ x,  ~ y, ~ .label,
  0.5,  1.1,  "bolditalic('Time '*t - 2)",
  1.5,  1.1,  "bolditalic('Time '*t - 1)",
  2.5,  1.1,  "bolditalic('Time '*t)"
)

# Adjust the length of the edges
complex_dag_contemp <- shorten_dag_arrows(
  complex_dag_contemp, 
  proportion = 0.06
  )

# Generate the DAG for the contemporaneous effect of X on Y
ggplot(
  data = complex_dag_contemp, 
  aes(x = x, y = y, xend = xend, yend = yend)
) +
  # Add the graph edges
  geom_dag_edges(
    aes(
      x = xstart, 
      y = ystart,
      edge_color = .edge_colour
      ), 
    edge_width = 1.5
  ) +
  # Add the graph nodes
  geom_dag_node(alpha = 0) +
  # Add the graph text
  geom_dag_text(
    aes(label = label),
    parse = TRUE,
    size = 11,
    color = "white",
    family = "serif",
    show.legend = FALSE
  ) +
  # Apply theme settings
  theme_dag(
    base_size = 24,
    base_family = "serif",
    strip.background = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(color = "white"),
    plot.subtitle = element_text(color = "white"),
    plot.caption = element_text(color = "white", face = "italic"),
    legend.title = element_text(color = "white", face = "bold"),
    legend.text = element_text(color = "white", face = "bold"),
    plot.caption.position = "plot",
    legend.position = "top"
  ) +
  # Add a legend for the edge colors
  scale_edge_color_identity(
    guide = "legend", 
    name = NULL,
    labels = c(
      "Treatment Path", 
      "Biasing Paths",
      "Other Paths"
    )
  ) +
  # Tweak the legend aesthetics
  guides(
    edge_alpha = "none",
    edge_color = "none") +
  # Plot Annotations
  annotate(
    "text",
    x = period_labels$x,
    y = period_labels$y,
    label = period_labels$.label,
    parse = TRUE,
    colour = "white",
    size = 12,
    family = "serif"
  )
```

## Licensing and Copyright Information

-   All text and images in this course are made available for public non-commercial use under a [Creative Commons Attribution-NonCommercial 4.0 International (CC BY-NC 4.0) License](https://creativecommons.org/licenses/by-nc/4.0/).

-   All R, HTML, and CSS code is provided for public use under a [BSD 3-Clause License](https://github.com/ajnafa/psci-3300-political-science-research/blob/main/LICENSE).

-   The files and code necessary to reproduce the content of this course are or will be made available via [the course's github repository](https://github.com/ajnafa/psci-3300-political-science-research) with the exception of those covered by existing commercial copyright restrictions (i.e., copies of the assigned readings for the course).

## References
