---
title: "Problem Set 0: Getting Started with R and Markdown"
author: "A. Jordan Nafa" ## Replace this with your Name
date: '`r format(Sys.time(), "%A, %B %d %Y")`' ## This code automatically generates the date
output: 
  html_document:
    highlight: haddock # You can also try tango, pygments, kate, monochrome, espresso, zenburn, haddock, breezedark, textmate, arrow, or rstudio here
    toc: true
    self_contained: true
    theme: journal # See here: https://www.datadreaming.org/post/r-markdown-theme-gallery/ for different theme options
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = " "
  )
```

# <!--# We can specify custom css formatting via <style></style> -->
<style>
body {
  text-align: justify;
  width: auto;
}
</style>

## Instructions

Welcome to PSCI 3300.003 Political Science Research Methods! RMarkdown allows us to create dynamically reproducible self-contained documents and reports. We'll use R and RMarkdown for problem sets throughout the semester and for your final projects in this course. In this class our uses will be fairly simple, but if you run into problems [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/) and the [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/) provide detailed book-length open-source resources for how to use RMarkdown.

The first problem set in this course is designed to ensure you have managed to successfully install the tools we'll be using throughout the semester. You’ll need to save the markdown file for the problem set to your computer so name it something like `lastname-firstname-problemset-0.Rmd` where `lastname` and `firstname` are your own last and first name respectively.

After following the instructions for installing [R](https://cloud.r-project.org/), [RStudio](https://www.rstudio.com/products/rstudio/), and [Rtools](https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html) (Windows) or [Xcode](https://apps.apple.com/us/app/xcode/id497799835?mt=12) (OSX) under the "Getting Started with R" module on the course's Canvas page, you should be able to open and compile this document in RStudio by clicking the **Knit** button. If everything is successfully configured, there should be a file called `lastname-firstname-problemset-0.html` in the same directory where you saved the first file to. **This is the file you need to Submit to Canvas via the "Problem Set 0: Getting Started with R and Markdown" assignment link, not the original .Rmd markdown file**. 

The due date for problem set 0 is Sunday, September 4th, 2022 at 11:59 PM.

# <!--# This is a code comment in html. It won't show up in the rendered html document but can be useful for clarifying technical details. -->

## Embedding R Code

We can execute R code within markdown documents as shown below. The code in this block will load the `iris` data from the base data sets package and print a summary of each column.

```{r}
# Load the iris data set
data(iris, package = "datasets")

# Print a summary of each column
summary(iris)
```

## Including Plots

We can easily embed plots as well. Here we'll use the base `plot` function to create a simple scatter plot depicting the relationship between two columns of the iris data.

```{r, echo=FALSE}
plot(
  x = iris$Sepal.Length,
  y = iris$Petal.Length, 
  pch = 21,
  bg = "blue",
  xlab = "Sepal Length", 
  ylab = "Petal Length"
  )
```

Notice that this returned the plot but didn't include the the code chunk. This is because we set `echo=FALSE` in the chunk header which prevents the code from printing but still returns the result. Try setting `echo=TRUE` in the code above and see what changes when you render the document.
