---
title: "Introduction to Quantitative Political Research"
author: "A. Jordan Nafa"
date: "August 30th, 2022"
institute: "University of North Texas"
subtitle: "PSCI 3300.003 Political Science Research Methods"
format: 
  revealjs:
    theme: blood
    highlight-style: monokai
    toc-depth: 2
    self-contained: true
    citations-hover: true
    code-link: true
    code-block-bg: "#272822"
    css: "../../../../assets/slide-style.css"
editor: visual
width: 1360
height: 800
bibliography: "../../../../assets/references.bib"
csl: "../../../../assets/apsa.csl"
link-citations: yes
---

## Introduction {#sec-introduction}

::: incremental
-   **Quantitative Social Science** is the use of quantitative data to study, analyze, or predict social and political phenomena

-   Outside of academic settings, this is more commonly known as *data science*

    -   The logic of quantitative social inquiry we will cover in this class is applied in academic social sciences, government, non-profits, and in the private sector

-   This course introduces students to the logic of causal inference and the tools quantitative social scientists use to study social, political, and economic phenomena
:::

## Course Staff

::: incremental
-   Instructor: A. Jordan Nafa

    -   Bachelor of Science in Government, Texas Woman's University (2014-2018)

    -   PhD Candidate, UNT Department of Political Science (2018-Present)

    -   Expertise in applied Bayesian statistics, causal inference, and women in politics

-   Teaching Assistant: Sung Min Yun

    -   PhD Student, UNT Department of Political Science (2021-Present)
:::

## Office Hours and Communication

::: incremental
-   Instructor: A. Jordan Nafa ([adamnafa\@my.unt.edu](adamnafa@my.unt.edu))

    -   Office: Wooten Hall 171

    -   Office Hours: Tuesday and Thursday 12:30-2:00 PM or by Appointment

-   Teaching Assistant: Sung Min Yun ([sungminyun\@my.unt.edu](sungminyun@my.unt.edu))

    -   Office: Wooten Hall 174A

    -   Office Hours: Wednesday 2:00-5:00 PM

-   We will try to respond to emails within 48 hours Monday-Friday

    -   See the syllabus for details on the policy about not responding to emails asking questions that are answered by the syllabus
:::

## Why Take this Course?

::: incremental
-   PSCI 3300 is a required course for political science majors so you'll have to take it at some point

-   The skills taught in this course are standard across an increasing number of industries and fields of research

    -   A basic knowledge of statistics, programming, and quantitative reasoning prepares you for a range of future career opportunities or graduate school

-   Contemporary political science is a primarily quantitative field, so understanding the tools of the trade will help you understand and critique the things you read in your other classes

-   No exams!
:::

## Teaching Philosophy

::: incremental
-   Math for the sake of math is an unfortunately common but largely unproductive way to teach quantitative social science

-   A college-level course in fundamentals of computer programming or elementary statistics may be helpful but is not required

-   I assume you can add, subtract, multiply, divide, and have a basic understanding of order of operations--I am sure some of you will prove me wrong

-   Our focus in this course will be primarily on how to apply, interpret, and evaluate analyses in political science rather than on the statistical theory behind them

-   This course is taught from a primarily Bayesian perspective but we'll cover some recent developments in frequentist estimation approaches as well
:::

## Teaching Philosophy

-   Though there is some introductory statistics required, in practice that means less of this $$\text{Normal}(y|\mu,\sigma) = \frac{1}{\sqrt{2 \pi}\, \sigma} \exp\left( - \, \frac{1}{2}\left(\frac{y -\mu}{\sigma} \right)^2 \right)$$

-   And more hands-on stuff like this

```{r, echo=TRUE}
## Simulate 10,000 random draws from a standard normal dist
std_norm <- rnorm(n = 10e3, mean = 0, sd = 1)

## Print a summary
summary(std_norm)
```

## Structure of the Course

::: incremental
-   Part I: Fundamentals of Political Research

    -   Research questions, surveying literature, basic logic of cause and effect, theory building, and measurement

-   Part II: Foundations of Quantitative Social Science

    -   Foundations of applied statistics, probability theory, Bayesian inference, and applied regression

-   Part III: Research Design and Causal Inference

    -   Experimental design, hypothesis testing and theory evaluation, strategies for causal inference with observational data, and the limits of magical thinking
:::

## Required Course Materials

::: incremental
-   There is one required textbook for this course which can be purchased from the UNT campus bookstore or Amazon

    -   ::: {style="color: #00A950"}
        Bueno de Mesquita, Ethan, and Anthony Fowler. *Thinking Clearly with Data: A Guide to Quantitative Reasoning and Analysis*. Princeton University Press, 2021. ISBN: 9780691214351
        :::

-   Although it is not required, I also recommend obtaining a copy of

    -   ::: {style="color: #00A950"}
        Andrew Gelman, Jennifer Hill, and Aki Vehtari. *Regression and Other Stories*. Cambridge University Press, 2021. ISBN: 9781107676510
        :::

-   All additional readings for the course are provided via Canvas
:::

## Technical Requirements

::: incremental
-   To complete the requirements for this course you will need access to a laptop or desktop computer with a stable internet connection

-   If you do not have access to a personal computer that meets the minimum requirements necessary to run R, you can access both R and RStudio from computers on the UNT campus

-   The computers in the political science lab in Wooten Hall 173 should have the most recent version of R, RStudio, and RTools

-   If you expect to have difficulty meeting the technical requirements for this course you need to let me know immediately
:::

## Class Meetings

::: incremental
-   Tuesday/Thursday from 9:30-10:50 AM

-   Part of our course time will be spent on lecture and answering questions and the other half will be spent working with data in R

-   You should come to class having done the readings on the syllabus for that day so we can discuss anything that isn't clear or appears confusing

    -   This is not a course you can skate through without ever reading anything and if you try to do this you will very likely fail.

    -   Advice on managing reading loads in the next class on Thursday
:::

## Problem Sets

-   You will complete five problem sets which will account for 40% of your overall grade and ask you to apply the things we cover in class using real or simulated data

```{r}
    ## Tibble of the assignemnts and Due Dates
    tibble::tribble(
      ~ assignment, ~ topic, ~ date,
      "0", "Getting Started with R and Markdown", "September 5th",
      "1", "Data Wrangling and Visualization", "September 25th",
      "2", "Probability and Descriptive Statistics", "October 9th",
      "3", "Regression and Inferential Statistics", "October 30th",
      "4", "Observational Causal Inference", "November 20th"
    ) |>
      ## Kable to create a table
      kableExtra::kbl(
        align = "ccc",
        row.names = FALSE,
        col.names = c("Assignment", "Topic", "Due Date")
      ) |>
      kableExtra::kable_material()
```

## Final Project

::: incremental
-   Develop a research question relevant to political science or public policy, identify the necessary source(s) of data, and undertake an analysis using R to answer the question

-   You have the option to work in groups of up to 4 people but you must provide written notice with the names of each student in the group **no later than Friday, September 23rd**

-   Broken down into a series of smaller assignments over the course of the semester

    -   Examples of what each of these are expected to look like and document templates will be provided via Canvas

-   Accounts for 50% of your overall grade; 70% from 8-12 page research paper and 30% from a final presentation
:::

## Course Participation

::: incremental
-   Remaining 10% of your grade is course participation

-   Showing up to class, asking relevant questions, and completing occasional in-class activities

-   Gives me wiggle room to pad grades at the end of the semester

-   Might also have in-class extra credit opportunities on days when attendance is scarce, but no promises
:::

## Getting Started with R

::: incremental
-   Detailed instructions for installing R, RTools/Xcode, RStudio, and configuring Stan are provided on the course's Canvas page

-   We will also make use of the probabilistic programming language *Stan* via it's R interfaces `brms` and `rstanarm` during parts of this course

-   If you have difficulty getting these installed on your personal computer please come by one of our office hours or make an appointment for assistance

-   First problem set due **Monday, September 5th** to verify you have successfully completed these installations
:::

## What Should You do This Week?

::: incremental
-   Read the full syllabus for more details about the course and important deadlines

-   Follow the instructions on Canvas for installing R, RStudio, and Stan

    -   We will go over any issues you have getting these installed on Thursday

-   Start thinking about an *empirical* political research question (More on this next class)
:::

## Licensing and Copyright Information

-   All text and images in this course are made available for public non-commercial use under a [Creative Commons Attribution-NonCommercial 4.0 International (CC BY-NC 4.0) License](https://creativecommons.org/licenses/by-nc/4.0/).

-   All R, HTML, and CSS code is provided for public use under a [BSD 3-Clause License](https://github.com/ajnafa/psci-3300-political-science-research/blob/main/LICENSE).

-   The files and code necessary to reproduce the content of this course are or will be made available via [the course's github repository](https://github.com/ajnafa/psci-3300-political-science-research) with the exception of those covered by existing commercial copyright restrictions (i.e., copies of the assigned readings for the course).
