---
title: 
  course: "Introduction to Quantitative Political Research"
  lecture: "PSCI 3300 Political Science Research Methods"
author: "A. Jordan Nafa"
date: "August 30th, 2022"
output: 
  beamer_presentation:
    slide_level: 3
    highlight: haddock
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: "../../../../assets/slide-template.tex"
    incremental: true
header-includes:
- \setbeamertemplate{itemize item}{\color{UNTGreenS1}$\blacktriangleright$}
- \setbeamertemplate{itemize subitem}{\color{UNTGreenS2}$\triangleright$}
bibliography: "../../../../assets/references.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  comment = " "
  )
```


## Introduction 

- **Quantitative Social Science** is the use of quantitative data to study, analyze, or predict social and political phenomena

-  Outside of academic settings, this is more commonly known as *data science*

   - The logic of quantitative social inquiry we will cover in this class is applied in academic social sciences, government, non-profits, and in the private sector
  
- This course introduces students to the logic of causal inference and the tools social scientists use to study social, political, and economic phenomena
  
  - We will focus primarily on quantitative approaches to the study of politics with an emphasis on application, causal reasoning, and prediction

## Why Take this Course?

-  PSCI 3300 is a required course for political science majors so you'll have to take it at some point

- The skills taught in this course are standard across an increasing number of industries and fields of research

  - A basic knowledge of statistics, programming, and quantitative reasoning prepares you for a range of future career opportunities or graduate school
  
  - Contemporary political science is a primarily quantitative field, so understanding the tools of the trade will help you understand and critique the things you read in your other classes
  
- I will not require you to laugh at my jokes

## Teaching Philosophy

- Math for the sake of math is an unfortunately common but largely unproductive way to teach quantitative social science

- A college-level course in fundamentals of computer programming or elementary statistics may be helpful but is not required

- Our focus in this course will be primarily on how to apply, interpret, and evaluate analyses in political science rather than on the statistical theory behind them

- Course is taught primarily from a pragmatic Bayesian perspective

- Though there is some basic statistics required, that means less of this $$\text{Normal}(y|\mu,\sigma) = \frac{1}{\sqrt{2 \pi}\, \sigma} \exp\left( - \, \frac{1}{2}\left(\frac{y -\mu}{\sigma} \right)^2 \right)$$

---

### Teaching Philosophy

- And more hands-on stuff like this

```{r}
## Simulate 10,000 random draws from a standard normal dist
std_norm <- rnorm(n = 10e3, mean = 0, sd = 1)

## Print a summary
summary(std_norm)
```

## Required Course Materials

- There is one required textbook for this course which can be purchased from the UNT campus bookstore or Amazon
  
  - \fullcite{BuenodeMesquita2021}
  
- Although it is not required, I also recommend obtaining a copy of

  - \fullcite{Gelman2021}

-  All additional readings for the course and various instructional resources for applied statistics and programming in R are provided via the course’s Canvas page

### Technical Requirements

- To complete the requirements for this course you will need access to a laptop or desktop computer with a stable internet connection

- If you do not have access to a personal computer that meets the minimum requirements necessary to run R, you can access both R and RStudio from computers on the UNT campus 
  
- The computers in the political science lab in Wooten Hall 173 should have the most recent version of R, RStudio, and RTools
  
- If you expect to have difficulty meeting the technical requirements for this course you need to let me know immediately

---

### Getting Started with R

- Detailed instructions for installing R, RTools/Xcode, RStudio, and configuring Stan are provided on the course's Canvas page

- We will also make use of the probabilistic programming language Stan via it's R interfaces `brms` and `rstanarm` during parts of this course

- If you have difficulty getting these installed on your personal computer please come by my office hours or make an appointment for assistance

## Course Structure

- **Class Meetings**

  - Tuesday/Thursday from 9:30 AM -- 10:50 AM
  
  - Part of our course time will be spent on lecture and answering questions and the other half will be spent working with data in R
  
  - You should come to class having done the readings on the syllabus for that day so that we can discuss anything that isn't clear or appears confusing
  
  - 10\% of your grade in this course consists of participation. This means showing up to class, asking relevant questions, and completing occasional in-class activities

---

### Course Structure

- **Problem Sets**

  - You will complete five problem sets which will account for 40\% of your overall grade and ask you to apply the things we cover in class using real or simulated data

  - Instructions for each problem set will be posted on the course's Canvas page approximately two weeks before its respective due date
  
\footnotesize
\begin{tabular}[t]{ccc}
\toprule
Assignment & Topic & Due Date\\
\midrule
Problem Set 0 & Getting Started with R and Markdown & September 4th, 2022\\
Problem Set 1 & Data Wrangling and Visualization & September 25th, 2022\\
Problem Set 2 & Probability and Descriptive Statistics & October 9th, 2022\\
Problem Set 3 & Regression and Inferential Statistics & October 30th, 2022\\
Problem Set 4 & Regression for Causal Inference & November 20th, 2022\\
\bottomrule
\end{tabular}
  
---

### Course Structure

- **Final Project**

  - Develop a research question relevant to political science or public policy, identify the necessary source(s) of data, and undertake an analysis using R to answer the question
  
  - You have the option to work in groups of up to 4 people but you must provide written notice with the names of each student in the group **no later than Friday, September 23rd**
  
  - Broken down into a series of smaller assignments over the course of the semester
  
  - Accounts for 50\% of your overall grade; 70\% from 8-12 page research paper and 30\% from a final presentation

## What Should You do Today?

- Read the full syllabus for more details about the course and important deadlines

- Follow the instructions on Canvas for installing R, RStudio, and Stan

- Start thinking about an empirical political research question