---
title: "Causal Relationships in the Quantitative Social Sciences I"
author: "A. Jordan Nafa"
date: "September 13th, 2022"
institute: "University of North Texas"
subtitle: "PSCI 3300.003 Political Science Research Methods"
format: 
  revealjs:
    theme: blood
    highlight-style: monokai
    toc-depth: 2
    self-contained: true
    citations-hover: true
    code-link: true
    transition: "fade"
    code-block-bg: "#272822"
    logo: "../../../../inst/figures/psci-3300-dag-trimmed-hex.png"
    css: "../../../../assets/slide-style.css"
    include-in-header: "../../../../assets/math-colors.js"
linkcolor: "#FF3030"
editor: visual
width: 1360
height: 800
bibliography: "../../../../assets/references.bib"
csl: "../../../../assets/apsa.csl"
link-citations: yes
---

```{r, include=FALSE}
## Load required packages
pacman::p_load(
  "tidyverse",
  "vdemdata", 
  install = FALSE
  )

## Base theme for the figures
fig_theme <- theme_light(base_size = 24, base_family = "serif") +
  ## Settings sepcific to the reveal.js theme
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(color = "white"),
    plot.subtitle = element_text(color = "white"),
    plot.caption = element_text(color = "white", face = "italic"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white", face = "bold"),
    legend.title = element_text(color = "white", face = "bold"),
    legend.text = element_text(color = "white", face = "bold"),
    strip.text = element_text(face = "bold"),
    plot.caption.position = "plot",
    legend.position = "top"
  )
```

## Overview

::: incremental
-   Better understanding of correlation and causation

-   Ways of thinking about and formally expressing causal relationships

-   No class Thursday, I will be at a conference in Canada

-   Research Question Assignment is due Sunday, September 18th

-   Problem set two will be posted on Canvas by the end of the day today
:::

$$
\definecolor{treat}{RGB}{27,208,213}
\definecolor{outcome}{RGB}{98,252,107}
\definecolor{baseconf}{RGB}{244,199,58}
\definecolor{covariates}{RGB}{178,26,1}
\definecolor{index}{RGB}{37,236,167}
\definecolor{timeid}{RGB}{244,101,22}
\definecolor{mu}{RGB}{71,119,239}
\definecolor{sigma}{RGB}{219,58,7}
\newcommand{normalcolor}{\color{white}}
\newcommand{treat}[1]{\color{treat} #1 \normalcolor}
\newcommand{resp}[1]{\color{outcome} #1 \normalcolor}
\newcommand{sample}[1]{\color{baseconf} #1 \normalcolor}
\newcommand{covar}[1]{\color{covariates} #1 \normalcolor}
\newcommand{obs}[1]{\color{index} #1 \normalcolor}
\newcommand{tim}[1]{\color{timeid} #1 \normalcolor}
\newcommand{mean}[1]{\color{mu} #1 \normalcolor}
\newcommand{vari}[1]{\color{sigma} #1 \normalcolor}
$$

## What is Correlation?

**Correlation** is the degree to which two or more features of the world tend to occur in tandem.

::: incremental
-   We'll call these features of the world "variables"

-   If two variables $\treat{X}$ and $\resp{Y}$ tend to occur together or increase at the same rate, we would say they are **positively correlated**

-   If the occurrence of $\treat{X}$ unrelated to $\resp{Y}$, we would say these two variables are **uncorrelated**

-   If when $\treat{X}$ occurs we are less likely to observe $\resp{Y}$, we would say these two variables are **negatively correlated**
:::

## The Resource Curse

Consider the case of the resource curse in comparative politics--an alleged negative correlation between dependence on oil production and democracy

|               | Not a Major Oil Producer | Major Oil Producer |
|---------------|:------------------------:|:------------------:|
| Democracy     |            78            |         15         |
| Non-Democracy |            58            |         26         |

-   Probability that a country is a democracy and a major oil producer

    -   $\Pr(\mathrm{Democracy} | \mathrm{No~Oil}) = \frac{78}{78+58} \approx 0.5735$

    -   $\Pr(\mathrm{Democracy} | \mathrm{Oil}) = \frac{15}{15+26} \approx 0.3659$

## Example: The Resource Curse

Consider the case of the resource curse in comparative politics--an alleged negative correlation between dependence on oil production and democracy

|                   | Not a Major Oil Producer | Major Oil Producer | Pr(Oil) |
|-------------------|:------------------------:|:------------------:|:-------:|
| **Democracy**     |            78            |         15         | 0.1613  |
| **Autocracy**     |            58            |         26         | 0.3095  |
| **Pr(Democracy)** |          0.5735          |       0.3659       |         |

-   Oil producing countries are less likely to be democracies than non-oil producing countries. Why?

## Democracy and Gender Equality

```{r continuous-correlation, echo=FALSE, dpi=300, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center', cache=TRUE}
gender_data <- vdem %>% 
  # Use just the year 2018
  filter(year == 2018) %>%  
  # Transmute a subset of the data for plotting
  select(
    country_name,
    v2x_polyarchy, 
    v2x_gender
    )

# Initiate the plot object
ggplot(gender_data, aes(x = v2x_gender, y = v2x_polyarchy)) +
  # Add the data points
  geom_point(
    aes(fill = country_name), 
    show.legend = FALSE, 
    shape = 21,
    size = 3
    ) +
  # Color for the fill
  scale_fill_viridis_d() +
  # Labels for the plot
  labs(
    x = "Women's Political Empowerment Index",
    y = "Procedural Democracy"
  ) +
  # Adjust the x axis scales
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Adjust the y axis scales
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Add annotation
  annotate(
    "text",
    x = 0.3,
    y = 0.7,
    size = 12,
    color = "white",
    family = "serif",
    label = latex2exp::TeX(r'($\rho_{x,y} = 0.7954$)'),
    parse = TRUE
  ) +
  # Plot theme settings
  fig_theme
```

## What is Correlation Good For?

::: incremental
- **Description**

  - Suppose we want to know whether countries where gender equality is higher are more democratic on average
  
  - We might be interested in the correlation between **gender equality** and **democracy**
  
  - If our data is good, we can provide a yes or no answer to this question with few assumptions needed
:::

## What is Correlation Good For?

We can estimate this correlation in R with a simple linear model and data from the Varieties of Democracy project's `{vdemdata}` package

```{r, echo=TRUE, eval=FALSE}
# Load the tidyverse library
pacman::p_load(
  "tidyverse",
  "latex2exp"
  )

# Install the vdemdata package if you didn't read the R programming tutorial
remotes::install_github("vdeminstitute/vdemdata")
```

## What is Correlation Good For?

We can estimate this correlation in R with a simple linear model and data from the Varieties of Democracy project's `{vdemdata}` package

```{r, echo=TRUE}
# Get the data we need from the vdemdata package
vdem_df <- vdemdata::vdem %>% 
  # We'll use just the year 2018 here for simplicity
  filter(year == 2018) %>%  
  # Transmute a subset of the data for plotting
  transmute(
    country_name, 
    v2x_polyarchy = v2x_polyarchy*10, 
    v2x_gender = v2x_gender*10
    )

# Estimate the linear relationship
lm_democ_gender <- lm(v2x_polyarchy ~ v2x_gender, data = vdem_df)

# Print a summary of the result
broom::tidy(lm_democ_gender)
```

## Democracy and Gender Equality

```{r continuous-correlation-fit, echo=FALSE, dpi=300, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center', cache=TRUE}
# Initiate the plot object
ggplot(vdem_df, aes(x = v2x_gender, y = v2x_polyarchy)) +
  # Add the data points
  geom_point(
    aes(fill = country_name), 
    show.legend = FALSE, 
    shape = 21,
    size = 3
    ) +
  # Color for the fill
  scale_fill_manual(values = viridisLite::turbo(n = length(vdem_df$country_name))) +
  # Labels for the plot
  labs(
    x = "Women's Political Empowerment Index",
    y = "Procedural Democracy",
    title = latex2exp::TeX(r'($E\[Y_{i} | X_{i}\] = -3.03 + 1.12\times X_{i}$ or $E\[Y_{i} | X_{i}\] = \alpha + \beta_{1}X_{i}$)')
  ) +
  # Adjust the x axis scales
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Adjust the y axis scales
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Plot the linear regression line
  geom_abline(
    intercept = -3.03, 
    slope = 1.12,
    color = "white",
    lty = 2,
    size = 2
    ) +
  # Plot theme settings
  fig_theme
```

## What is Correlation Good For?

::: incremental
- **Prediction and Forecasting**

  - Suppose we want to predict which financial transactions are likely to be fraudulent
  
  - With large amounts of data on legitimate consumer transactions, we could develop a model that detects anomalies that are likely to be fraudulent
  
  - Major financial firms have entire teams dedicated to predicting and preventing fraud in this manner
  
  - Other examples of prediction include election forecasting, self-driving cars, and many other tasks
:::

## What is Correlation Good For?

::: incremental
- **Causal Inference**

  - Suppose we want to know if high school students would be more successful in life if they were forced to take calculus
  
  - The observed correlation between calculus and future success might be useful
  
  - But we would have to assume that the students taking calculus are otherwise the same as everyone else in terms of their underlying chances of success
  
  - Aside from very special circumstances, this kind of assumption will be hard to defend
  
  - As a general rule, correlation does not imply causation
:::

## Descriptive Statistics

For every variable we observe, we can compute a number of different statistics. Three that are particularly useful for understanding data are **mean**, **variance**, and **standard deviation**

::: incremental
- Mean $\mean{\mu}_{\treat{x}}$: $\frac{\sum_{\obs{i}=\obs{1}}^{\sample{n}} \treat{x}_{\obs{i}}}{\sample{N}}$

- Variance $\vari{\sigma}_{\treat{x}}^{2}$: $\frac{\sum_{\obs{i}=\obs{1}}^{\sample{n}} (\treat{x}_{\obs{i}} - \mean{\mu}_{\treat{x}})^{2}}{\sample{N}}$

- Standard Deviation $\vari{\sigma}_{\treat{x}}$: $\sqrt{\vari{\sigma}_{\treat{x}}^{2}}$

- The mean and variance are known as the first and second moments of a variable's distribution
::: 

## Univariate Descriptive Statistics in R

Continuing with our previous example, we can calculate the mean, standard deviation, and variance of a numeric variable in R

## Univariate Descriptive Statistics in R

Continuing with our previous example, we can calculate the mean, standard deviation, and variance of a numeric variable in R

```{r, echo=TRUE}
# Mean of women's political empowerment in 2018
mean(vdem_df$v2x_gender, na.rm = TRUE)
```

## Univariate Descriptive Statistics in R

Continuing with our previous example, we can calculate the mean, standard deviation, and variance of a numeric variable in R

```{r, echo=TRUE}
# Mean of women's political empowerment in 2018
mean(vdem_df$v2x_gender, na.rm = TRUE)

# Standard deviation of women's political empowerment in 2018
sd(vdem_df$v2x_gender, na.rm = TRUE)
```

## Univariate Descriptive Statistics in R

Continuing with our previous example, we can calculate the mean, standard deviation, and variance of a numeric variable in R

```{r, echo=TRUE}
# Mean of women's political empowerment in 2018
mean(vdem_df$v2x_gender, na.rm = TRUE)

# Standard deviation of women's political empowerment in 2018
sd(vdem_df$v2x_gender, na.rm = TRUE)

# Variance of women's political empowerment in 2018
var(vdem_df$v2x_gender, na.rm = TRUE)
```

## Univariate Descriptive Statistics in R

Continuing with our previous example, we can calculate the mean, standard deviation, and variance of a numeric variable in R

```{r, echo=TRUE}
# Mean of women's political empowerment in 2018
mean(vdem_df$v2x_gender, na.rm = TRUE)

# Standard deviation of women's political empowerment in 2018
sd(vdem_df$v2x_gender, na.rm = TRUE)

# Variance of women's political empowerment in 2018
var(vdem_df$v2x_gender, na.rm = TRUE)
```

These are univariate statistics, meaning they describe the distribution of a single variable

## Bivariate Descriptive Statistics

Now that we’ve learned some notation, and we can compute useful statistics for a single variable, we can start thinking about how to measure the correlation between two variables

::: incremental
- One useful measure of correlation is the **covariance**

  - $Cov_{\treat{x}, \resp{y}} = \frac{\sum_{\obs{i}=\obs{1}}^{\sample{n}} (\treat{x}_{\obs{i}} - \mean{\mu}_{\treat{x}})(\resp{y}_{\obs{i}} - \mean{\mu}_{\resp{y}})}{\sample{N}}$
  
- A second is the **correlation coefficient**

  - $\rho_{\treat{x}, \resp{y}} = \frac{Cov_{\treat{x}, \resp{y}}}{\vari{\sigma}_{\treat{x}} \cdot \vari{\sigma}_{\resp{y}}}$
::: 

## Bivariate Descriptive Statistics in R

We can calculate the covariance and correlation coefficient for two numeric variables in R

## Bivariate Descriptive Statistics in R

We can calculate the covariance and correlation coefficient for two numeric variables in R

```{r, echo=TRUE}
# Covariance for gender and democracy
cov(x = vdem_df$v2x_gender, y = vdem_df$v2x_polyarchy)
```

## Bivariate Descriptive Statistics in R

We can calculate the covariance and correlation coefficient for two numeric variables in R

```{r, echo=TRUE}
# Covariance for gender and democracy
cov(x = vdem_df$v2x_gender, y = vdem_df$v2x_polyarchy)

# Correlation coefficient for gender and democracy
cor(x = vdem_df$v2x_gender, y = vdem_df$v2x_polyarchy)
```

## Bivariate Descriptive Statistics in R

We can calculate the covariance and correlation coefficient for two numeric variables in R

```{r, echo=TRUE}
# Covariance for gender and democracy
cov(x = vdem_df$v2x_gender, y = vdem_df$v2x_polyarchy)

# Correlation coefficient for gender and democracy
cor(x = vdem_df$v2x_gender, y = vdem_df$v2x_polyarchy)
```

## Bivariate Descriptive Statistics in R

We could calculate the correlation and covariance manually as well for illustrative purposes

```{r, echo=TRUE}
# (x_{i} - mu_{x})
cov_x <- vdem_df$v2x_gender - mean(vdem_df$v2x_gender)
```

## Bivariate Descriptive Statistics in R

We could calculate the correlation and covariance manually as well for illustrative purposes

```{r, echo=TRUE}
# (x_{i} - mu_{x})
cov_x <- vdem_df$v2x_gender - mean(vdem_df$v2x_gender)

# (y_{i} - mu_{y})
cov_y <- vdem_df$v2x_polyarchy - mean(vdem_df$v2x_polyarchy)
```

## Bivariate Descriptive Statistics in R

We could calculate the correlation and covariance manually as well for illustrative purposes

```{r, echo=TRUE}
# (x_{i} - mu_{x})
cov_x <- vdem_df$v2x_gender - mean(vdem_df$v2x_gender)

# (y_{i} - mu_{y})
cov_y <- vdem_df$v2x_polyarchy - mean(vdem_df$v2x_polyarchy)

# (x_{i} - mu_{x})(y_{i} - mu_{y})
cov_xy <- cov_x*cov_y
```

## Bivariate Descriptive Statistics in R

We could calculate the correlation and covariance manually as well for illustrative purposes

```{r, echo=TRUE}
# (x_{i} - mu_{x})
cov_x <- vdem_df$v2x_gender - mean(vdem_df$v2x_gender)

# (y_{i} - mu_{y})
cov_y <- vdem_df$v2x_polyarchy - mean(vdem_df$v2x_polyarchy)

# (x_{i} - mu_{x})(y_{i} - mu_{y})
cov_xy <- cov_x*cov_y

# Sum of (x_{i} - mu_{x})(y_{i} - mu_{y})
sum_cov_xy <- sum(cov_xy)
```

## Bivariate Descriptive Statistics in R

We could calculate the correlation and covariance manually as well for illustrative purposes

```{r, echo=TRUE}
# (x_{i} - mu_{x})
cov_x <- vdem_df$v2x_gender - mean(vdem_df$v2x_gender)

# (y_{i} - mu_{y})
cov_y <- vdem_df$v2x_polyarchy - mean(vdem_df$v2x_polyarchy)

# (x_{i} - mu_{x})(y_{i} - mu_{y})
cov_xy <- cov_x*cov_y

# Sum of (x_{i} - mu_{x})(y_{i} - mu_{y})
sum_cov_xy <- sum(cov_xy)

# Covariance of x and y
(cov_result <- sum_cov_xy/length(cov_xy))
```

## Bivariate Descriptive Statistics in R

We could calculate the correlation and covariance manually as well for illustrative purposes

```{r, echo=TRUE}
# (x_{i} - mu_{x})
cov_x <- vdem_df$v2x_gender - mean(vdem_df$v2x_gender)

# (y_{i} - mu_{y})
cov_y <- vdem_df$v2x_polyarchy - mean(vdem_df$v2x_polyarchy)

# (x_{i} - mu_{x})(y_{i} - mu_{y})
cov_xy <- cov_x*cov_y

# Sum of (x_{i} - mu_{x})(y_{i} - mu_{y})
sum_cov_xy <- sum(cov_xy)

# Covariance of x and y
(cov_result <- sum_cov_xy/length(cov_xy))

# sigma_{x} sigma_{y}
sigma_xy <- sd(vdem_df$v2x_gender)*(vdem_df$v2x_polyarchy)
```

## Bivariate Descriptive Statistics in R

We could calculate the correlation and covariance manually as well for illustrative purposes

```{r, echo=TRUE}
# (x_{i} - mu_{x})
cov_x <- vdem_df$v2x_gender - mean(vdem_df$v2x_gender)

# (y_{i} - mu_{y})
cov_y <- vdem_df$v2x_polyarchy - mean(vdem_df$v2x_polyarchy)

# (x_{i} - mu_{x})(y_{i} - mu_{y})
cov_xy <- cov_x*cov_y

# Sum of (x_{i} - mu_{x})(y_{i} - mu_{y})
sum_cov_xy <- sum(cov_xy)

# Covariance of x and y
(cov_result <- sum_cov_xy/length(cov_xy))

# sigma_{x} sigma_{y}
sigma_xy <- sd(vdem_df$v2x_gender)*(vdem_df$v2x_polyarchy)

# Correlation
cov_result/(sd(vdem_df$v2x_gender)*sd(vdem_df$v2x_polyarchy))
```

## Measuring Correlation

::: incremental
- The correlation coefficient tells us about the tightness of the relationship between two variables. 

- But we often care more about the substantive magnitude of the relationship. How much does $\resp{Y}$ vary as $\treat{X}$ varies?

- To answer this question, we want to know the **slope** of the regression line.

- $\beta = \frac{Cov_{\treat{x}, \resp{y}}}{\vari{\sigma}_{\treat{x}}^{2}}$

- On average, for every one-unit increase in $\treat{X}$, $\resp{Y}$ increases by...
:::

## Measuring Correlation

```{r linear-fit-example, echo=FALSE, dpi=300, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center'}
# Initiate the plot object
ggplot(vdem_df, aes(x = v2x_gender, y = v2x_polyarchy)) +
  # Add the data points
  geom_point(
    aes(fill = country_name), 
    show.legend = FALSE, 
    shape = 21,
    size = 3
    ) +
  # Color for the fill
  scale_fill_manual(values = viridisLite::turbo(n = length(vdem_df$country_name))) +
  # Labels for the plot
  labs(
    x = "Women's Political Empowerment Index",
    y = "Procedural Democracy",
    title = latex2exp::TeX(r'($E\[Y_{i} | X_{i}\] = -3.03 + 1.12\times X_{i}$ or $E\[Y_{i} | X_{i}\] = \alpha + \beta_{1}X_{i}$)')
  ) +
  # Adjust the x axis scales
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Adjust the y axis scales
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Plot the linear regression line
  geom_abline(
    intercept = -3.03, 
    slope = 1.12,
    color = "white",
    lty = 2,
    size = 2
    ) +
  # Plot theme settings
  fig_theme
```

## When $y = mx + b$ Fails

::: incremental
- An unfortunate fact of life, though one some social scientists all too often fail to appreciate, is that not everything is approximately linear and additive

- In cases of non-linear relationships, calculating the linear correlation between two variables will give us the wrong answer

  - **If you make stupid assumptions, you will get stupid results**
  
  - Causation need not imply **linear correlation**
  
- Lots of interesting relationships are non-linear and there are ways of analyzing these relationships

  - Pro-tip: The two most power tools for data analysis ever created are the histogram and the scatter plot
::: 

## Non-Linear Relationships

Easiest way to to illustrate the problem is usually through simulation

```{r, echo=TRUE}
# Simulate some non-linear relationships
nonlinear_sims <- tibble(
  x = runif(n = 2e3, min = -10, max = 10),
  y_posquad = x + x^2 + rnorm(2e3, 0, 2),
  y_negquad = x - x^2 + rnorm(2e3, 0, 2),
  y_sin = sin(x*3.14) + rnorm(2e3, 0, 0.5),
  y_linear = x + rnorm(2e3, 0, 2)
)
```

## Non-Linear Relationships

Easiest way to to illustrate the problem is usually through simulation

```{r, echo=TRUE}
# Simulate some non-linear relationships
nonlinear_sims <- tibble(
  x = runif(n = 2e3, min = -10, max = 10),
  y_posquad = x + x^2 + rnorm(2e3, 0, 2),
  y_negquad = x - x^2 + rnorm(2e3, 0, 2),
  y_sin = sin(x*3.14) + rnorm(2e3, 0, 0.5),
  y_linear = x + rnorm(2e3, 0, 2)
)
```

Then we can use `{ggplot2}` to make scatter plots of the relationships in the simulated variables

## Non-Linear Relationships

```{r echo=TRUE}
# Initiate the plot object
posquad_plot <- ggplot(nonlinear_sims, aes(x = x, y = y_posquad, fill = x)) +
  # Add the data points
  geom_point(shape = 21, size = 3) +
  # Add the linear fit
  geom_smooth(method = "lm", size = 2, se = FALSE, lty = 2, color = "white") +
  # Tweak the fill color scheme
  scale_fill_viridis_c() +
  # Labels for the plot
  labs(
    x = "X",
    y = "Y",
    title = latex2exp::TeX(r'($Y_{i} = X_{i} + X_{i}^{2} + \epsilon_{i}$)')
  ) +
  # Adjust the x axis scales
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Adjust the y axis scales
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8))
```

## Non-Linear Relationships

```{r quadratic-example, echo=FALSE, dpi=300, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center'}
# Print the plot object
posquad_plot +
  # Plot theme settings
  fig_theme +
  # Disable legends
  guides(fill = "none")
```

## Non-Linear Relationships

```{r echo=TRUE}
# Initiate the plot object
negquad_plot <- ggplot(nonlinear_sims, aes(x = x, y = y_negquad, fill = x)) +
  # Add the data points
  geom_point(shape = 21, size = 3) +
  # Add the linear fit
  geom_smooth(method = "lm", size = 2, se = FALSE, lty = 2, color = "white") +
  # Tweak the fill color scheme
  scale_fill_viridis_c() +
  # Labels for the plot
  labs(
    x = "X",
    y = "Y",
    title = latex2exp::TeX(r'($Y_{i} = X_{i} - X_{i}^{2} + \epsilon_{i}$)')
  ) +
  # Adjust the x axis scales
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Adjust the y axis scales
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8))
```

## Non-Linear Relationships

```{r negquadratic-example, echo=FALSE, dpi=300, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center'}
# Print the plot object
negquad_plot +
  # Plot theme settings
  fig_theme +
  # Disable legends
  guides(fill = "none")
```

## Non-Linear Relationships

```{r echo=TRUE}
# Initiate the plot object
sin_plot <- ggplot(nonlinear_sims, aes(x = x, y = y_sin, fill = x)) +
  # Add the data points
  geom_point(shape = 21, size = 3) +
  # Add the linear fit
  geom_smooth(method = "lm", size = 2, se = FALSE, lty = 2, color = "white") +
  # Tweak the fill color scheme
  scale_fill_viridis_c() +
  # Labels for the plot
  labs(
    x = "X",
    y = "Y",
    title = latex2exp::TeX(r'($Y_{i} = Sin(X_{i}\cdot \pi) + \epsilon_{i}$)')
  ) +
  # Adjust the x axis scales
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Adjust the y axis scales
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8))
```

## Non-Linear Relationships

```{r sin-example, echo=FALSE, dpi=300, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center'}
# Print the plot object
sin_plot +
  # Plot theme settings
  fig_theme +
  # Disable legends
  guides(fill = "none")
```

## Licensing and Copyright Information

-   All text and images in this course are made available for public non-commercial use under a [Creative Commons Attribution-NonCommercial 4.0 International (CC BY-NC 4.0) License](https://creativecommons.org/licenses/by-nc/4.0/).

-   All R, HTML, and CSS code is provided for public use under a [BSD 3-Clause License](https://github.com/ajnafa/psci-3300-political-science-research/blob/main/LICENSE).

-   The files and code necessary to reproduce the content of this course are or will be made available via [the course's github repository](https://github.com/ajnafa/psci-3300-political-science-research) with the exception of those covered by existing commercial copyright restrictions (i.e., copies of the assigned readings for the course).

## References
