---
title: "Introduction to Quantitative Political Research"
author: "A. Jordan Nafa"
date: "September 1st, 2022"
institute: "University of North Texas"
subtitle: "PSCI 3300.003 Political Science Research Methods"
format: 
  revealjs:
    theme: blood
    highlight-style: monokai
    toc-depth: 2
    self-contained: true
    citations-hover: true
    code-link: true
    code-block-bg: "#272822"
    css: "../../../../assets/slide-style.css"
editor: visual
width: 1360
height: 800
bibliography: "../../../../assets/references.bib"
csl: "../../../../assets/apsa.csl"
link-citations: yes
---

```{r, include=FALSE}
## Load required packages
pacman::p_load(
  "tidyverse",
  install = FALSE
  )

## Base theme for the figures
fig_theme <- theme_light(base_size = 24, base_family = "serif") +
  ## Settings sepcific to the reveal.js theme
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(color = "white"),
    plot.subtitle = element_text(color = "white"),
    plot.caption = element_text(color = "white", face = "italic"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white", face = "bold"),
    legend.title = element_text(color = "white", face = "bold"),
    legend.text = element_text(color = "white", face = "bold"),
    strip.text = element_text(face = "bold"),
    plot.caption.position = "plot",
    legend.position = "top"
  )
```

## How Should we Study Politics? {#sec-introduction}

Political science is a field characterized by a diverse range of approaches to inquiry and debates about how we ought to study political phenomena have long animated the discipline.

::: incremental
-   **Normative**

    -   Focused on subjective, moral questions about the world and how things *ought to be*.

-   **Empirical**

    -   Focused on objective explanation and description, questions about how the world *is* and *why.*

-   **Interpretive**

    -   
:::

## Normative Approaches

Normative approaches to the study of politics date back thousands of years and feature prominently in the sub-field of political philosophy.

::: incremental
-   How should the world look? Asks for a moral judgement

    -   Who should be responsible for paying for the consequences of climate change?

    -   Would the world be better off if we fired Elon Musk into the Sun?

    -   Should women have autonomy over their reproductive choices?

-   These types of questions may indeed be important, but they do not lend themselves to scientific answers

-   Normative arguments are common in certain areas of law and philosophy but have little place in this course
:::

## Empirical Approaches

Empirical approaches are those that aim to apply the scientific method to the study of politics and hold a dominant place contemporary political science.

::: incremental
-   Empirical approaches can be descriptive or causal, quantitative or qualitative, experimental or observational but they all aim to answer some question about *how, what, or why the world is*.

-   **Description** focuses on observing and measuring the state of the world; it aims to answer questions about *who* or *what* in relation to some phenomena [@Gerring2012].

    -   What is democracy and how can we operationalize it?

    -   Who won the 2020 presidential election election?
:::

------------------------------------------------------------------------

### 2020 Presidential Election Vote Totals {.smaller}

```{r, echo=FALSE, dpi=300, dev.args = list(bg = 'transparent'), fig.height=9, fig.width=16, fig.align='center'}
## 2020 election totals from the federal election commission
fec_totals <- tribble(
  ~ candidate, ~ popvote, ~ ecvote, ~ votepct,
  "Joe Biden", 81268924/1000000, 306, 51.3,
  "Donald Trump", 74216154/1000000, 232, 46.9
) %>% 
  ## Pivot to long form
  pivot_longer(cols = popvote:ecvote)

## Define labels for the facets
facet_labs <- as_labeller(
  x = c(
    "popvote" = "Popular Vote (in Millions)",
    "ecvote" = "Electoral College Vote"
  )
)

## Make a plot of the results
ggplot(fec_totals, aes(y = value, x = candidate, fill = candidate)) +
  ## Facet by popular vote/EC vote
  facet_wrap(~ name, scales = "free_y", labeller = facet_labs) +
  ## Bar plot geom
  geom_bar(stat = "identity", color = "black") +
  ## Change the colors of the fill
  scale_fill_manual(values = c("firebrick", "royalblue2")) +
  ## Adjust the lower bound of the y axis
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 10), 
    expand = expansion(0.01, 0.01)
    ) +
  ## Add labels to the plot
  labs(
    fill = "2020 Candidate",
    x = "",
    y = "Vote Total",
    caption = "Source: United States Federal Election Commission, 2020"
  ) +
  fig_theme
```

## Empirical Approaches

::: incremental
-   **Causal** approaches are concerned with explaining *why* some phenomenon occurs in the world [@Samii2016].

-   Contemporary political science is a discipline interested in answering *causal questions*.

    -   Why do poor conservatives tend to vote against their economic interests?

    -   Do gender-inclusive peace processes reduce the risk of conflict recurrence?

    -   How do campaign contributions influence legislative behavior?

-   Our focus in this class will be primarily on causal questions and entirely on empirical approaches to the study of politics
:::

## References
